How to build an homogeneous language for heterogeneous platforms

### 2013-06-06 papers/gpce2013/gpce2013.tex
Progress on sections 2 and 3. Still no title.

\begin{itemize}
 \item we define high-level abstractions typically used in Web programming generating low-level code leveraging
native Web browsers APIs;
 \item in the case of abstractions shared between servers and clients, we specialize the code generation in order to
leverage the target platform native environments.
\end{itemize}

We validate our approach with a case study implemented with various candidate technologies and discuss the relative
pro and cons of them. Though the code written in our DSL is high-level and can be shared between clients and servers,
it has the same runtime performances as hand-tune low-level JavaScript code.

@@

Lightweight Modular Staging~\cite{Rompf12_LMSThesis} is a framework for defining deeply embedded DSLs in Scala using
staging~\cite{Jorring1986_Staging}. It has been used to define high-performance DSLs for parallel
computing~\cite{Brown11_Parallel}.

@@

Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
example, consider listing \ref{null-unsafe} finding a particular widget in the page and then a particular
button in the widget. The native \code{querySelector} method returns \code{null} if no node matched the given
selector in the document. If we run this code in a page where the widget is not present, it will throw an error
and stop further JavaScript execution. Defensive code can handle \code{null} reference, but leads to very cumbersome
code, as shown in listing \ref{null-defensive}.

@@

Listing \ref{null-js-scala} implements in \emph{js-scala} a program equivalent to listing \ref{null-defensive}. The
\code{for} notation is used to sequence computations over optional values encoded with a monad, allowing a high
level of expressiveness. (To be continued, mention \code{Rep[Option[A]]}, \code{Rep[Some[A]]}, \code{Rep[None.type]})

We want to define a DSL that has both the safety and performance of listing \ref{null-defensive} but the
expressiveness of listing \ref{null-unsafe}. We can get safety by wrapping potentially null values of type
\code{Rep[A]} in a container of type \code{Rep[Option[A]]} requiring explicit dereferencing, we can get
expressiveness by using the Scala \code{for} notation for dereferencing, and finally we can get performance by
generating code that does not actually wraps values in a container but instead checks if they are \code{null} or not
when dereferenced. The wrapping container exists only at the stage-level and is removed during the code generation.
Listing \ref{null-js-scala} shows how to write the same program using our DSL (implementation details are
given in section \ref{implementation}). The evaluation of this listing produces a graph of statements from which
JavaScript code equivalent to listing \ref{null-defensive} is generated.

@@

@@

@@

\begin{figure}
\begin{lstlisting}[language=HTML,caption=HTML,label=html]
<div class=article>
  <span>French wine: </span>
  <strong>10</strong>
</div>
\end{lstlisting}
\end{figure}

In this section we show how we can define a template engine as an embedded DSL with minimal effort. This template
engine is statically typed and able to insert dynamic content in a safe way. It provides a powerful expression
language, requires no extra compilation step and can be used on both client-side and server-side.

Because the template engine is defined as en embedded DSL, we can reuse Scala’s constructs:

\begin{itemize}
\item a function taking some parameters and returning a DOM fragment directly models a template taking parameters and
returning a DOM fragment~;
\item the type system type-checks template definitions and template calls~;
\item the Scala language itself is the expression language~;
\item compiling a template is the same as compiling user code.
\end{itemize}

So the only remaining work consists in defining the DSL vocabulary to define DOM nodes. We provide a \code{el}
function to define a tag and any \code{String} value is considered to be a text node.

@@

Listing \ref{forest} uses our DSL and generates a code equivalent to listing \ref{dom-api}. The readability has
been highly improved: nesting tags is just like nesting code blocks, HTML entities are
automatically escaped in text nodes, developers have the full computational power of Scala to inject dynamic data and
DOM fragments definitions are written using functions so they compose just as functions compose. These benefits come
with no performance loss because the DSL generates code building DOM fragments by using the native JavaScript API.

Our DSL is equivalent to a template engine with Scala as the expression language. Making it usable on both server and
client sides was surprisingly as simple as defining another code generator for the DSL, producing Scala code.

For instance, the template written in listing \ref{forest} produces the following Scala code usable on
server-side:

\begin{lstlisting}
def articleUi(article: Article) = {
  val x0 = <span>{ article.name + ": " }</span>
  val x1 = <strong>{ article.price }</strong>
  val x2 =
    <div class="article">
      {List(x0, x1)}
    </div>
  x2
}
\end{lstlisting}

We are able to tackle the code sharing issues described in the introduction (more details?) because of the embdedded
nature of our DSLs: dynamic content of templates is written using embedded DSLs too, so their translation into
JavaScript and Scala is managed by their respective code generators.

>>>>

\begin{itemize}
 \item we define high-level abstractions typically used in Web programming, that generate low-level code leveraging
native Web browsers APIs;
 \item in the case of abstractions shared between servers and clients, we specialize the code generation in order to
leverage the target platform native environments.
\end{itemize}

We validate our approach with a case study implemented with various candidate technologies and discuss the relative
pro and cons of them. Though the code written in our DSL is high-level and can be shared between clients and servers,
it has the same runtime performances as hand-tuned low-level JavaScript code.

@@

Lightweight Modular Staging~\cite{Rompf12_LMSThesis} (LMS) is a framework for defining deeply embedded DSLs in Scala
using staging~\cite{Jorring1986_Staging}. It has been used to define high-performance DSLs for parallel
computing~\cite{Brown11_Parallel}. (Todo. Mention \code{Rep[_]})

@@

Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
example, consider listing \ref{null-unsafe} finding a particular widget in the page and then a particular
button in the widget. The native \code{querySelector} method returns \code{null} if no node matched the given
selector in the document. If we run this code in a page where the widget is not present, it will throw an error
and stop further JavaScript execution. Defensive code can be written to handle \code{null} references, but leads to
very cumbersome code, as shown in listing \ref{null-defensive}.

@@

Listing \ref{null-js-scala} implements in \emph{js-scala} a program equivalent to listing \ref{null-defensive}. The
\code{for} notation is used to sequence computations over optional values encoded with a monad. The \code{find}
function returns a \code{Rep[Option[Element]]} value, that can either be a \code{Rep[Some[Element]]} (if an element
was found) or a \code{Rep[None.type]} (if no element was found). The \code{for} expression contains a sequence
of statements that are executed in order, as long as the previous statement returned a \code{Rep[Some[Element]]}
value.

@@

Such a monadic API brings both safety and expressiveness to developers manipulating optional values but usually
involves the creation of an extra container object holding the optional value. In our case, we expose
the monadic API at the first stage but we generate code that does not wrap values in a container object but instead
checks if they are \code{null} or not when dereferenced. So the extra container object exists only at the
stage-level and is removed during code generation: listing \ref{null-js-scala} produces a code equivalent to listing
\ref{null-defensive}.

Last but not least, we leverage the modular architecture of LMS to define two code generators for this API: one
targeting JavaScript and the other targeting Scala. So the same abstraction is efficiently translated on both server
and client sides.

@@

This section shows how we define an abstraction shared between clients and servers, as in the previous section, but
that has different native counterparts on client and server sides. The challenge is to define an API providing a
common vocabulary that generates code using the target platform native APIs.

@@

\begin{figure}
\begin{lstlisting}[caption=Scala XML API,label=scala-xml-api]
def articleUi(article: Article) =
  <div class="article">
    <span>{ article.name + ": " }</span>
    <strong>{ article.price }</strong>
  </div>
\end{lstlisting}
\end{figure}

A common task in Web applications consists in computing HTML fragments representing a part of the page content. This
task can be performed either from the server-side (to initially respond to a request) or from the client-side (to
update the current page). As an example, listing \ref{dom-api} defines a JavaScript function \code{articleUi} that
builds a DOM tree containing an article description, and listing \ref{scala-xml-api} shows how one could implement
a similar function on server-side using the standard Scala XML library. The reader may notice that the client-side
and server-side API are very different and that the client-side API is very low-level and inconvenient to use.

Instead, we provide a common high-level DSL for defining HTML fragments and we make this DSL generate code leveraging
native environments. Listing \ref{forest} shows how to implement our example with our DSL. The \code{el} function
defines an HTML element, eventually containing attributes and children elements. Any children of an element that is
not an element itself is converted into a text node. The children elements of an element can also be obtained
dynamically from a collection, as shown in listing \ref{forest-loops}.

@@

\begin{figure}
\begin{lstlisting}[label=forest-loops,caption=Using loops]
def articlesUi(articles: Rep[Seq[Article]]) =
    el('ul)(
        for (article <- articles)
        yield el('li)(articleUi(article))
    )
\end{lstlisting}
\end{figure}

The \code{el} function returns an \code{Element} IR node that is a tree composed of other \code{Element} nodes and
\code{Text} nodes. This tree is traversed by the code generators to produce code building an equivalent DOM tree on
client-side and code building an equivalent XML fragment on server-side. When the children elements of an element are
constant values (as in listing \ref{forest}) rather than dynamically obtained (as in listing \ref{forest-loops}), the
code generators inline the loop that adds children to their parent, for better performance. As a result, listing
\ref{forest} generates a code equivalent to listing \ref{dom-api} on client-side and equivalent to
\ref{scala-xml-api} on server-side.
